---
title: "coartic_midpts"
author: "Meg Cychosz"
date: "3/30/2019"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include=FALSE}
library('tidyverse')
library('dplyr')
library('ggplot2')
library('lme4')
library('lmerTest')
```

An analysis of coarticulation across versus within morphemes in adult and child Quechua. Here we are analyzing adjacent coarticulation between [a] and [p] across morphemes as in "llama-pi" and within morphemes as in "api". 

# Load data
```{r import code, warning=FALSE, message=FALSE}
data <- read.csv('all_files_mfccs_3tpt.csv', na.strings = c("NA","NaN", " "))
```

```{r pre-cleaning, echo=FALSE}
data <- data %>%
  filter(Note!='DON\'T USE')
data$Note <- droplevels(data$Note)

data <- data %>%
  filter(Phone!='t\'')
data <- data %>%
  filter(Phone!='k\'')
data$Phone <- droplevels(data$Phone)
```

# Prepare and process data
```{r create df, echo=FALSE}
morph_df <- data[ which(data$Word=='api-pi' | data$Word=='imilla-pi' | 
                       data$Word=='juk\'ucha-pi' | data$Word=='llama-pi' |
                       data$Word=='papa-pi' | data$Word=='sunkha-pi' |
                       data$Word=='t\'ika-pi' | data$Word=='uhut\'a-pi' |
                       data$Word=='waka-pi' | data$Word=='wallpa-pi' | 
                       data$Word=='wawa-pi' |  data$Word == 'api' |
                         data$Word == 'papa'), ]

imilla <- morph_df[ which(morph_df$Word=='imilla-pi' & (morph_df$Phone=='a' | morph_df$Phone == 'p')), ]
jukucha <- morph_df[ which(morph_df$Word=='juk\'ucha-pi' & (morph_df$Phone=='a' | morph_df$Phone == 'p')), ]
llama <- morph_df[ which((morph_df$Word=='llama-pi' & (morph_df$Phone=='a' & morph_df$Previous == 'm')) |
                        ((morph_df$Word=='llama-pi' & morph_df$Phone == 'p'))), ]
sunkha <- morph_df[ which(morph_df$Word=='sunkha-pi' & (morph_df$Phone=='a' | morph_df$Phone == 'p')), ]
tika <- morph_df[ which(morph_df$Word=='t\'ika-pi' & (morph_df$Phone=='a' | morph_df$Phone == 'p')), ]
uhuta <- morph_df[ which(morph_df$Word=='uhut\'a-pi' & (morph_df$Phone=='a' | morph_df$Phone == 'p')), ]
wawa <- morph_df[ which((morph_df$Word=='wawa-pi' & (morph_df$Phone=='a' & morph_df$Following == 'p')) |
                           ((morph_df$Word=='wawa-pi' & morph_df$Phone == 'p'))), ]
waka <- morph_df[ which((morph_df$Word=='waka-pi' & (morph_df$Phone=='a' & morph_df$Following == 'p')) |
                          ((morph_df$Word=='waka-pi' & morph_df$Phone == 'p'))), ]
wallpa <- morph_df[ which((morph_df$Word=='wallpa-pi' & (morph_df$Phone=='a' & morph_df$Previous == 'p')) |
                      ((morph_df$Word=='wallpa-pi' & morph_df$Following=='i'))), ]

# get 'papa'
a <-
  morph_df[which(morph_df$Morph_status == 'root' &
                   morph_df$Phone == 'a'),]
# get the first 'a' that matches this criteria bc we're looking morpheme-internal:
a.new <-
  a[seq(1, nrow(a), 2),] # index every second row, starting at row 1, to get 1st 'a' meas
p <-
  morph_df[which(morph_df$Morph_status == 'root' &
                   (morph_df$Word == 'papa-pi')),]
p.new <- p[which(p$Previous == 'a'),]

# get 'ap from papa
ap <- morph_df[which(morph_df$Morph_status == 'morph'),]
papa2 <- rbind(a.new, p.new, ap)

# get api
api <-
  morph_df[which(morph_df$Word == 'api' &
                   (morph_df$Phone == 'a' | morph_df$Previous == 'a')),]
api_pi <-
  morph_df[which(morph_df$Word == 'api-pi' &
                   (morph_df$Phone == 'a' | morph_df$Previous == 'a')),]
api2 <- rbind(api,api_pi)
```

```{r, concatenate word_df}
new_morphDF <- Reduce(function(x,y) merge(x,y,all=TRUE) ,list(api2, imilla,
            jukucha, llama, papa2, sunkha, tika, uhuta, waka, wallpa, wawa)) 
new_morphDF$Word <- droplevels(new_morphDF$Word)
```

\newpage
# Convert structure of spectral measurements
```{r pre-process, echo=TRUE}
new_morphDF$Spectrum_int_2 <- gsub( ']', '', new_morphDF$Spectrum_int_2)
new_morphDF$Spectrum_int_2 <- gsub( '[ ', '', new_morphDF$Spectrum_int_2, fixed = TRUE)

# make string, then numeric
new_morphDF$variable_sep_tp2 <- str_extract_all(new_morphDF$Spectrum_int_2, "[-0-9\\.]+")
new_morphDF$spec_vector_tp2 <- lapply(new_morphDF$variable_sep_tp2 , FUN = as.numeric)
new_morphDF <- as.data.frame(new_morphDF)
```

# Calculate euc distance between vectors
```{r plyr_check}
# sanity check
if(any(grepl("package:plyr", search()))) detach("package:plyr") else message("plyr not loaded")
```

```{r calculations on spec vectors}
# euclidean distance and raw difference
diff_morphdf <- new_morphDF %>% 
  group_by(Word, Speaker, Note) %>% 
  mutate(raw_diff = map2(spec_vector_tp2, lead(spec_vector_tp2), `-`)) %>%  # sanity check
  mutate(euc_dist = map2(spec_vector_tp2, lead(spec_vector_tp2), function(x, y) 
    sqrt(sum((x-y) ^ 2)))) %>% 
  as.data.frame()  

# sequence duration
diff_morphdf <- diff_morphdf %>% 
  group_by(Word, Speaker, Note) %>% 
  mutate(ap_duration = map2(Phone_duration, lead(Phone_duration), `+`)) %>% 
  as.data.frame()  

# remove NA rows where measurement was made upon but not stored
df.final <- subset(diff_morphdf, euc_dist != 'NA')
df.final$euc_dist <- as.numeric(df.final$euc_dist)
df.final$ap_duration <- as.numeric(df.final$ap_duration)
```

\newpage
# Descriptive stats
```{r rename variables, echo=FALSE, warning=FALSE, message=FALSE}
library('plyr')
df.final$Age_group <- mapvalues(df.final$Speaker, from = 
            c('Alfredo', 'Andrea', 'Brian', 'Cristian', 'Cristina', 'Cynthia', 'Daisy',
              'Daniela', "David", "Diego", "Dismal", "Edgar", "Eduard", "Elias", "Elsa",
              "Eusevia", "Evan", "Fabiola", "Gladys", "Ivan", "Johan", "Josefina",
              "Joselin", "Juana", "Justina", "Laura", "Leonilda", "Liber", "Lucia",
              "LuisMario", "Margarita", "Mari", "Martha", "Nayeli", "Rosemary", "Senovia",
              "Sulma", "Sulmaadult", "Vanesa", "Vladimir"), 
            to = c("9", "adult", "8", "7", "9", "5-6", "7", 
                   "8", "5-6", "8", "10", "9", "7", "7", "adult", 
                   "adult", "8", "7", "adult", "5-6", "5-6", "adult",
                   "10", "adult", "7", "adult", "7", "7", "5-6",
                   "8", "10", "8", "adult", "9", "9", "adult",
                   "10", "adult", "10", "7"))

df.final$Age <- mapvalues(df.final$Age_group, from = 
                            c('adult', '9', '8', '7', '5-6', '10'), 
                          to = c("adult", "child", "child", "child", 
                                 "child", "child"))

df.final$Morph_status <- mapvalues(df.final$Word, from = 
c('api-pi', 'imilla-pi', 'juk\'ucha-pi', 'llama-pi', 'papa-pi', 'uhut\'a-pi', 
't\'ika-pi', 'wallpa-pi', 'waka-pi', 'wawa-pi', 'sunkha-pi', 'api', 'papa'), 
to = c("within morpheme", "across morpheme", "across morpheme", "across morpheme", 
"within morpheme", "across morpheme", "across morpheme", "across morpheme", "across morpheme", "across morpheme", "across morpheme", "within morpheme", "within morpheme"))
detach(package:plyr)

adult <- subset(df.final, Age =='adult')
child <- subset(df.final, Age =='child')
ten <- subset(df.final, Age_group =='10')
nine <- subset(df.final, Age_group =='9')
et <- subset(df.final, Age_group =='8')
svn <- subset(df.final, Age_group =='7')
six <- subset(df.final, Age_group =='5-6')
```

### averages by age
```{r echo=FALSE}
print(paste("Average euc. distance between spectral vectors, all environments: adults",mean(adult$euc_dist)%>% round(2)))
print(paste("Average euc. distance, all environments: children",mean(child$euc_dist)%>% round(2)))
print(paste("Average euc. distance, all environments: ten y/os",mean(ten$euc_dist) %>% round(2)))
print(paste("Average euc. distance, all environments: nine y/os",mean(nine$euc_dist)%>% round(2)))
print(paste("Average euc. distance, all environments: eight y/os",mean(et$euc_dist)%>% round(2)))
print(paste("Average euc. distance, all environments: seven y/os",mean(svn$euc_dist)%>% round(2)))
print(paste("Average euc. distance, all environments: 5 & 6 y/os",mean(six$euc_dist)%>% round(2)))
```

### age*morphological status
```{r, echo=FALSE}
a.suffix <- subset(df.final, Morph_status == 'across morpheme' & Age == 'adult')
a.root <- subset(df.final, Morph_status == 'within morpheme' & Age == 'adult')
c.suffix <- subset(df.final, Morph_status == 'across morpheme' & Age == 'child')
c.root <- subset(df.final, Morph_status == 'within morpheme' & Age == 'child')
```

```{r, echo=FALSE}
print(paste("Average euc. distance across morphemes: adults",mean(a.suffix$euc_dist)%>% round(2)))
print(paste("Average euc. distance within morphemes: adults",mean(a.root$euc_dist)%>% round(2)))
print(paste("Average euc. distance across morphemes: children",mean(c.suffix$euc_dist)%>% round(2)))
print(paste("Average euc. distance within morphemes: children",mean(c.root$euc_dist)%>% round(2)))
```

\newpage
# visualize some stuff
```{r chil-adult visual, echo=TRUE, warning=FALSE}
jpeg('age_eucdist.jpeg', width = 500, height=500)

ggplot(df.final, aes(Age, euc_dist, fill = Morph_status)) +
  geom_boxplot() + guides(fill = guide_legend(title = "Position")) +
  ylim(4, 30) +
  scale_fill_manual(values=c("gray39", "gray77")) + 
  labs(title = "Spectral distance between \n middle thirds of [a] & [p]",
            y = "Euclidean distance", x = "Age") +
  theme(legend.position = c(0.6, 0.88)) +
  theme(legend.title=element_text(size=12,face="bold")) +
  theme(legend.text=element_text(size=12)) +
  theme(plot.title = element_text(size = 18, face = "bold")) +
  theme(
    axis.text = element_text(face = "bold", size = 22),
    axis.title = element_text(size=22,face="bold"))

dev.off()
```

\newpage
### make new age groups
```{r by age group, warning=FALSE}
library('plyr')
df.final$Age_group <- mapvalues(df.final$Age_group, from = 
                            c('adult', '9', '10', '8', '7', '5-6'), 
                          to = c("adult", "9-10", "9-10", "7-8", 
                                 "7-8", "5-6"))

df.final$Age_group <- factor(df.final$Age_group,
                             levels = c("5-6", "7-8", "9-10", "adult"))

jpeg('agegroup_eucdist.jpeg', width = 500, height=500)

ggplot(df.final, aes(x = Age_group, y = euc_dist, fill = Morph_status)) +
  geom_boxplot() + guides(fill = guide_legend(title = "Position")) + 
  scale_fill_manual(values=c("gray39", "gray77")) + 
  ylim(4, 30) +
  labs(title = "Spectral distance between \n middle thirds of [a] & [p]",
            y = "Euclidean distance", x = "Age (in years)") +
  theme(legend.position = c(0.69, 0.88)) +
  theme(legend.title=element_text(size=12,face="bold")) +
  theme(legend.text=element_text(size=12)) +
  theme(plot.title = element_text(size = 18, face = "bold")) +
  theme(
    axis.text = element_text(face = "bold", size = 22),
    axis.title = element_text(size=22,face="bold"))

dev.off()
```

\newpage
# visualize distance by word
```{r, warning=FALSE}
adult <- subset(df.final, Age == 'adult')
child <- subset(df.final, Age == 'child')

ggplot(child, aes(x = Word, y = euc_dist, fill = Morph_status)) +
  geom_boxplot() +
  guides(fill = guide_legend(title = "Word")) +
  ylim(4,30) +
  labs(title = "Spectral distance between \n middle third of [a] & [p] by word in children",
            y = "Euclidean distance", x = "Age") +
  theme(axis.title = element_text(size = 16)) +
  theme(plot.title = element_text(size = 18, face = "bold")) +
  theme(
    axis.text.x = element_text(face = "bold", size = 6),
    axis.text.y = element_text(face = 'bold', size = 12)
  )

ggplot(adult, aes(x = Word, y = euc_dist, fill = Morph_status)) +
  geom_boxplot() + guides(fill = guide_legend(title = "Word")) +
  ylim(4,30) +
  labs(title = "Spectral distance between \n middle third of [a] & [p] by word in adults",
            y = "Euclidean distance", x = "Age") +
  theme(axis.title = element_text(size = 16)) +
  theme(plot.title = element_text(size = 18, face = "bold")) +
  theme(
    axis.text.x = element_text(face = "bold", size = 6),
    axis.text.y = element_text(face = 'bold', size = 12)
  )
```


\newpage
# Fit some models

### baseline and baseline + ap_duration
Less coarticulation in temporally longer segments, as expected. 
```{r, message=FALSE}
summary(m <- lmer(euc_dist ~ + (1 | Speaker) + (1 | Word), data = df.final)) 
summary(mdur <- lmer(euc_dist ~ ap_duration + (1 | Speaker) + (1 | Word), data = df.final)) 
anova(m, mdur) # improves
```

### Morphological position + sequence_duration
```{r, message=FALSE}
summary(m1.a <- lmer(euc_dist ~ Morph_status + ap_duration + (1 | Speaker) + (1 | Word), data = df.final)) 
anova(mdur, m1.a) # doesn't improve 
```

\newpage
### Age[child v adult] + sequence_duration + Morph_status
```{r, message=FALSE}
summary(m1 <- lmer(euc_dist ~ Age + ap_duration + Morph_status + (1 | Speaker) + (1 | Word), data = df.final)) 
anova(mdur, m1) # doesn't improve
```

\newpage
### Age*Morphological status + sequence duration    

The interaction of age and morphological status improves model fit. Adults differentiate between morphological environments, but children do not. Specifically, adults coarticulate less within morphemes than across. 
```{r, message=FALSE}
summary(m2 <- lmer(euc_dist ~ Age*Morph_status + ap_duration + (1 | Speaker) + (1 | Word), data = df.final)) 
anova(mdur, m2) # improves
```



### Morphological status*sequence_duration

No reliable interaction of morphological status and sequence duration. 
```{r, message=FALSE}
summary(m4 <- lmer(euc_dist ~ Morph_status*ap_duration + (1 | Speaker) + (1 | Word), data = df.final)) 
anova(mdur, m4) # doesn't improve

```



# There is no relationship between segment duration and morphological environment - just less coarticulation in longer segments in all environments.
```{r duration*morph visual, warning=FALSE}
ggplot(df.final, aes(ap_duration*1000, euc_dist, color=Morph_status, fill=Morph_status)) +
 geom_point(size=.5) +
 geom_smooth(method = "lm") +
 labs(x = "[a-p] sequence duration (ms)", y = "Euclidean distance") +
 ggtitle("Coarticulation by [a-p] sequence duration \n and morphological environment") +
 ylim(4, 30)
```

\newpage
# Duration by morphological status by age

This model shows that there is a three-part interaction between morphological position, segment duration, and age. 
```{r, message=FALSE}
summary(m5 <- lmer(euc_dist ~ Morph_status*ap_duration*Age + (1 | Speaker) + (1 | Word), data = df.final)) 
anova(mdur, m5) # improves
```

\newpage
```{r, warning=FALSE}
#jpeg("interaction.jpg", width = 500, height = 500)

ggplot(df.final, aes(ap_duration*1000, euc_dist, color=Morph_status, fill=Morph_status)) +
 geom_point(size=.5) +
 geom_smooth(method = "lm") +
 labs(x = "[a-p] sequence Duration (ms)", y = "Euclidean distance") +
  facet_grid(~Age) +
 ggtitle("[a]-[p] coarticulation by sequence duration, \n morphological environment, and age") +
 ylim(4, 30)

#dev.off()
```

```{r effects by word, warning=FALSE}
adultdf <- df.final %>%
  filter(Age=='adult')
#jpeg("byword.jpg", width = 750, height = 750)

ggplot(adultdf, aes(ap_duration*1000, euc_dist, color=Morph_status, fill=Morph_status)) +
 geom_point(size=.5) +
 geom_smooth(method = "lm") +
 labs(x = "[a-p] sequence Duration (ms)", y = "Euclidean distance") +
  facet_wrap(~Word, nrow=4) +
 ggtitle("Coarticulation by sequence duration, word, \n and morphological environment in adult speakers") +
 ylim(4, 30)

#dev.off()
```

